---
title: "NEON forecasting challenge"
author: "rachel torres"
date: "2025-01-22"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(phenor)
library(lubridate)
```

## NEON phenology data viz 

[steps to code are from this webste](https://projects.ecoforecast.org/neon4cast-ci/targets.html)

Description of data:

**gcc_90** - P1D - Green chromatic coordinate is the ratio of the green digital number to the sum of the red, green, blue digital numbers from a digital camera.	
Horizon: 30 days	
Latency: ~ 2 days

**rcc_90** -	P1D	- Red chromatic coordinate is the ratio of the Red digital number to the sum of the red, green, blue digital numbers from a digital camera.	
Horizon: 30 days	
Latency: ~ 2 days


First, select a site where phenology data is available ([see table](https://projects.ecoforecast.org/neon4cast-ci/targets.html) where phenology is marked 1)

In example below I filter by ABBY (Abby Road in Oregon) and SJER (San Joaquin Experimental Range in California)

```{r data}

#set download url 
url <- "https://sdsc.osn.xsede.org/bio230014-bucket01/challenges/targets/project_id=neon4cast/duration=P1D/phenology-targets.csv.gz"

# get data frame from website 
phenology_targets <- read_csv(url, show_col_types = FALSE)
  
# filter by site id 
sites <- phenology_targets %>% filter(site_id=="ABBY" | site_id=="SJER")

# plot time series 
ggplot(sites) + 
  geom_point(aes(x=datetime, y=observation, col=site_id)) +
  facet_grid('variable')

```

Other ideas to explore in visualization:

- plots to make to explore - time series 
- smooth time series 
- monthly averages 
- monthly values with year by color 
- daily values by year 
- average daily values 

## Example Plot
```{r example}

# example plot that combines filter, mutating, aggregating 

sites %>% 
  # filter by site id 
  filter(site_id=="ABBY" & 
           # filter by variable - only gcc90
           variable=="gcc_90") %>%   
  # use lubridate functions to get different time steps
  mutate(month = month(datetime),
         year=year(datetime),
         yd=yday(datetime)) %>% 
  # summarize 
  group_by(yd, year) %>% 
  summarize(gcc90 = mean(observation)) %>% 
  # plot 
  ggplot(aes(x=yd, 
             y=gcc90, 
             col=as.factor(year))) + 
  geom_line()



```

Continue your own code below - what patterns can you find? 
