---
title: "NEON forecasting challenge"
author: "rachel torres"
date: "2025-01-22"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
# library(phenor) this is for later 
library(lubridate)
```

## NEON phenology data viz 

[steps to code are from this webste](https://projects.ecoforecast.org/neon4cast-ci/targets.html)

Description of data:

**gcc_90** - P1D - Green chromatic coordinate is the ratio of the green digital number to the sum of the red, green, blue digital numbers from a digital camera.	
Horizon: 30 days	
Latency: ~ 2 days

**rcc_90** -	P1D	- Red chromatic coordinate is the ratio of the Red digital number to the sum of the red, green, blue digital numbers from a digital camera.	
Horizon: 30 days	
Latency: ~ 2 days


First, select a site where phenology data is available ([see table](https://projects.ecoforecast.org/neon4cast-ci/targets.html) where phenology is marked 1)

In example below I filter by ABBY (Abby Road in Oregon) and SJER (San Joaquin Experimental Range in California)

```{r data}

#set download url 
url <- "https://sdsc.osn.xsede.org/bio230014-bucket01/challenges/targets/project_id=neon4cast/duration=P1D/phenology-targets.csv.gz"

# get data frame from website 
phenology_targets <- read_csv(url, show_col_types = FALSE)
  
# filter by site id 
sites <- phenology_targets %>% filter(site_id=="SJER")

# plot time series 
ggplot(sites) + 
  geom_point(aes(x=datetime, y=observation, col=site_id)) +
  facet_grid('variable')

```

Other ideas to explore in visualization:

- plots to make to explore - time series 
- smooth time series 
- monthly averages 
- monthly values with year by color 
- daily values by year 
- average daily values 

## Example Plot
```{r example}

# example plot that combines filter, mutating, aggregating 

sites %>% 
  # filter by site id 
  filter(site_id=="ABBY" & 
           # filter by variable - only gcc90
           variable=="gcc_90") %>%   
  # use lubridate functions to get different time steps
  mutate(month = month(datetime),
         year=year(datetime),
         yd=yday(datetime)) %>% 
  # summarize 
  group_by(yd, year) %>% 
  summarize(gcc90 = mean(observation)) %>% 
  # plot 
  ggplot(aes(x=yd, 
             y=gcc90, 
             col=as.factor(year))) + 
  geom_line()


# to use only 1 site, use the filter function 
# saving as new data frame 'site_dates'
site_dates = sites %>% filter(site_id=="SJER")
```

Continue your own code below - what patterns can you find? 

## Let's look at climate data for the NEON site

We are going to load more data using the `neonUtilities` R package - see examples [here](https://www.neonscience.org/resources/learning-hub/tutorials/neondatastackr)

We want to get air temperature values, which are available as 1 and 30 minute average from triple aspirated observations, and have an ID of DP1.00003.001.

See a full list of data products [here](https://data.neonscience.org/data-products/explore)

other IDs you might be interested 

- precipitation - DP1.00006.001
- shortwave radiation - DP1.00014.001
- photosynthetically active radiation (PAR) - DP1.00024.001

```{r neon_utils, eval=F}
# run this first to install 
#install.packages("neonUtilities")

# load package 
library(neonUtilities)

# get the start and end dates for the SJER site 
# first get the date column for only SJER 
site_dates = sites %>% filter(site_id=="SJER")
# the functions below select the first and last date entry in the datetime column, then format to Year-month 
start = format(site_dates$datetime[1], "%Y-%m")
end = format(site_dates$datetime[nrow(dates)], "%Y-%m")

# alternatively to above, you could hand select your start and end times, format has to be Year-month 
start = "2017-01"
end = "2025-01"

# download the triple aspirated temperature files - for site SJER only to start - check your console, this may take a minute 
# to download a different clim variable, change dpID
# triptemp <- loadByProduct(dpID=c("DP1.00003.001"),
#                           site=c("SJER"),
#                           startdate=start, 
#                           enddate=end)
# the format of this download is a list of 7 that contains metadata and data frames for 1 minute observations (TAAT_1min) and 30 minute (TAAT_30min)

phen <- loadByProduct(dpID=c("DP1.00033.001"),
                          site=c("SJER"),
                          startdate="2018-01", 
                          enddate="2018-12")

# get daily min, max, and mean temps 
minute_daily_avg = triptemp$TAAT_1min %>% 
  mutate(date = date(startDateTime)) %>% 
  group_by(date) %>% 
  summarize(tmin = min(tempTripleMean),
            tmax = max(tempTripleMean),
            tmean = mean(tempTripleMean)) 

ggplot(minute_daily_avg, 
       aes(x=date, y=tmean)) +
  geom_point() +
  geom_point(aes(x=date, y=tmin, col="tmin")) + 
  geom_point(aes(x=date, y=tmax, col="tmax"))

# save dataframe to computer so don't have to download again 
saveRDS(minute_daily_avg, "SJER_trip_temp_NEON.Rds")
```

## Get Accumulated growing degree days 
```{r agdd}

minute_daily_avg <- readRDS("../data/SJER_trip_temp_NEON.Rds")


# calculate gdd as the mean over the threshold (because we are working in celsius, the threshold is 0 or 32F, we only need the mean)

gdd = minute_daily_avg %>% 
  mutate(tmean = (tmin+tmax)/2,
         year = year(date),
         doy = yday(date)) %>% 
  select(year, doy, tmean) %>% 
  filter(year>2022) %>% 
  fill(tmean) # not the best strategy but will work for now - fills NA values with previous value 

# plot as avg temperature (gdd)
ggplot(gdd, aes(x=doy, y=tmean, col=as.factor(year))) +
  geom_point(alpha=0.5) + 
  geom_smooth() +
  labs(x="day of year", y="tmean (degC)", col="year") + 
  theme_bw()

# create a new column for cumulative values, starting at 0 for each year 
# cumsum is a base function that takes a vector and produces a new vector with cumulative values. try putting ?cumsum into your console to learn more, or visualize an example by putting `cumsum(c(1,2,3,4))` into your console
gdd <- gdd %>% 
  group_by(year) %>% 
  mutate(agdd = cumsum(tmean))

# plot agdd 
ggplot(gdd, aes(x=doy, y=agdd, col=as.factor(year))) + 
  geom_hline(aes(yintercept=1000), linetype="dashed") + 
  geom_line(linewidth=1) + 
  labs(x="day of year", y="AGDD (base=0C)", col="year") +
  theme_bw() + theme(legend.position = "inside",
                     legend.position.inside = c(0.15, 0.8))

```


## Combine phenology and climate data 

Continuing with temperature data, going to create a data frame that includes columns for dates, gcc90, and temperature, then will do example plot 

To do this, we will use a join function from the `dplyr` package. Read more [here](https://dplyr.tidyverse.org/reference/mutate-joins.html).

I am continuing to look at the SJER site only, where I downloaded the climate data. 

```{r joins}

# to join, we need one column to have the same identifier, in this case the date needs to be in the same format and column name 

# changing format of date in site_dates to match date in minute_daily_avg 
site_dates$date = ymd(site_dates$datetime)

# inner_join maintains all columns and joins by dates 
phen_temps = inner_join(site_dates, minute_daily_avg, by="date")

# preview new dataframe 
head(phen_temps)

summary(phen_temps)

# add other columns for different temporal groupings 
phen_temps <- phen_temps %>% 
  mutate(year=year(date),
         month=month(date),
         doy=yday(date))

# start thinking about how to plot to explore patterns 
phen_temps %>% 
  filter(variable=="gcc_90") %>% 
  ggplot(aes(x=tmin, y=observation, col=as.factor(month))) + geom_point() +
  labs(x="tmin (C)", y="gcc 90")

# compare gcc with spring temps 
phen_temps %>% 
  filter(variable=="gcc_90") %>%
  ggplot(aes(x=doy, y=observation, col=tmin)) + geom_point() +
  labs(x="day of year", y="gcc 90")
  

```

## Using a model 

fable and fable tools package

```{r fab}
# note: you may need to install before running library - uncomment below and run these lines 
# install.packages('fable')
# installed.packages('fabletools')

library(fable)
library(fabletools)

gcc_df <- phen_temps %>% 
   filter(variable=="gcc_90")

temp_lm <- tsibble(gcc_df, index='date') %>% 
  fabletools::model(
    m_tmin = fable::TSLM(observation ~ tmin),
    m_tmax = fable::TSLM(observation ~ tmax),
    m_t_doy = fable::TSLM(observation ~ tmin+doy)
  )

fabletools::report(temp_lm) %>%
  select(`.model`, AIC)
# lowest value is the model with temperature only but it is very small difference 


## ----plot tslm modeled vs observed, fig.height = 6, fig.width = 12, fig.align = "center", fig.cap="Figure: TSLM predictions of beetle abundances at OSBS compared against observed data"----
# visualize model fit
# augment reformats model output into a tsibble for easier plotting
fabletools::augment(temp_lm) %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = observation, lty = "Obs"), color = "dark gray") +
  geom_line(aes(y = .fitted, color = .model, lty = "Model")) +
  facet_grid(.model ~ .) +
  theme_bw()

# Load necessary libraries
library(dplyr)
library(ggplot2)
library(fable)
library(fabletools)
library(tsibble)

# Assuming your data frames are named 'eco_data' (ecological observations) 
# and 'climate_data' (climate observations), and both have a 'date' column

# 1. Join the two data frames by date
merged_data <- inner_join(site_dates, minute_daily_avg, by = "date")

# 2. Filter to only include rows where the variable is "gcc_90"
gcc_data <- merged_data %>%
  filter(variable == "gcc_90")

# 3. Generate a scatterplot of gcc_90 vs. temperature (assuming temperature column is 'temp')
ggplot(gcc_data, aes(x = tmin, y = observation)) +
  geom_point() +
  labs(title = "Scatterplot of GCC_90 vs. Temperature",
       x = "Temperature",
       y = "GCC_90 Observation") +
  theme_minimal()

# 4. Convert the filtered data into a tsibble for time series analysis
gcc_ts <- gcc_data %>%
  as_tsibble(index = date) %>% 
  mutate(doy=yday(date))

# Run a time series linear regression (TSLM) of gcc_90 observations against temperature
gcc_lm <- gcc_ts %>%
  model(TSLM(observation ~ tmin+doy))


# Print the summary of the regression model
report(gcc_lm)



```


