---
title: "pheno"
author: "rachel torres"
date: "2025-01-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install.packages("devtools")
#if(!require(remotes)){install.packages("remotes")}
#remotes::install_github("bluegreen-labs/phenor@v1.3.1")
#install.packages("phenocamr")
#install.packages("maps")

library(phenor)
library(phenocamr)
library(maps)
library(raster)
library(tidyverse)
```

## tutorial - NEON pheno packages 

https://www.neonscience.org/resources/learning-hub/tutorials/phenocam-phenor-modeling

For the latest version, install `phenor` from here: https://github.com/bluegreen-labs/phenor (be sure to update all packages for installation to work, and install GenSA)


```{r download}

# example from website: Bartlett NH 
# phenocamr::download_phenocam(
#   frequency = 3,
#   veg_type = "DB", #DB for decid broadlead, EN for evergreen, default is ALL 
#   roi_id = 1000,
#   site = "bartlettir",
#   phenophase = TRUE,
#   out_dir = "." # saves in current working dir
#   )

# evergreens in del norte county (close to north humboldt)
# phenocamr::download_phenocam(
#   frequency = 3,
#   veg_type = "EN", #DB for decid broadlead, EN for evergreen, default is ALL 
#   roi_id = 1000,
#   site = "delnortecounty1",
#   phenophase = TRUE,
#   out_dir = "." # saves in current working dir
#   )

# shrubs in crownpoint NM, NE of Gallup at Navajo Technical University 
#phenocamr::download_phenocam(
#  frequency = 3,
#  roi_id = 1000,
#  site = "ntupinyongarden",
#  phenophase = TRUE,
#  out_dir = "." # saves in current working dir
#  )

# two files: time series and transition data 
```
# explore data

Find the metadata here:
https://phenocam.nau.edu/webcam/tools/summary_file_format/


## use vignette 

https://bluegreen-labs.github.io/phenocamr/articles/phenocamr-vignette.html

source has clear steps, but uses base plotting 

```{r phenocamr}
# read as phenocamr type 
df = read_phenocam("example_data/bartlettir_DB_1000_3day.csv")

df <- expand_phenocam(df) # expand every 3 day to every day 

df <- detect_outliers(df) # detect outliers

df <- smooth_ts(df) # smooth data using AIC 

phenology_dates <- phenophases(df, internal = TRUE) # calculate rising and falling

as.data.frame(df$data) %>% 
  mutate(date = as.Date(date)) %>% 
  ggplot(aes(x=date, y=smooth_gcc_90)) + 
  geom_line() +
  labs(x="Date", y="GCC") + 
  # rising "spring" greenup dates 
  geom_vline(data=phenology_dates$rising, aes(xintercept=transition_50), col="green") + 
  # falling 'autumn' senescence dates 
  geom_vline(data=phenology_dates$falling, aes(xintercept=transition_50), col="brown")


```

## climate data: Daymet  

Use the function `merge_daymet`

Daymet is continuous, gridded estimates of weather and climate variables creted through statistical interpolation of ground based measurements. Read more here: https://daymet.ornl.gov/

```{r daymet}

bnh_clim = merge_daymet("example_data/bartlettir_DB_1000_3day.csv")

names(bnh_clim)

bnh_data = bnh_clim$data

# plot daily rainfall and temperatures 
ggplot(bnh_data, aes(x=date, y=prcp..mm.day.))+
  geom_line() + 
  labs(y="precip [mm]", title="Daily Precipitation") +
  theme_bw() 

ggplot(bnh_data) +
  geom_point(aes(x=date, y=tmax..deg.c., col="tmax"), col="red") +
  geom_point(aes(x=date, y=tmin..deg.c., col="tmin"), col="blue") +
  labs(y="temp. deg C", title = "Daily temperature") +
  theme_bw()

```

## Explore data with climate 

```{r explore_clim}

# day of the year patterns 
ggplot(bnh_data) +
  geom_point(aes(x=doy, y=gcc_90))

# precipitation
ggplot(bnh_data) + 
  geom_point(aes(x=prcp..mm.day., y=gcc_90, col=doy)) + scale_color_viridis_c()

# day of the year with daylength 
ggplot(bnh_data) + 
   geom_point(aes(x=doy, y=gcc_90, col=dayl..s.)) + scale_color_viridis_c()

# daylength 
ggplot(bnh_data) +
  geom_point(aes(x=doy, y=gcc_90, col=dayl..s.)) + scale_color_viridis_c()

# min temperatures 
ggplot(bnh_data) +
  geom_point(aes(x=tmin..deg.c., y=gcc_90))
# max temperatures 
ggplot(bnh_data) +
  geom_point(aes(x=tmax..deg.c., y=gcc_90))

```

## start with linear model 
```{r lm}

# subset data 
data_mod = bnh_data %>% dplyr::filter(year<2014)

# start with one variable 
lm_test = lm(data=data_mod, gcc_90~dayl..s.)
# see residuals, p-vals, Rsquares, etc. 
summary(lm_test)

# check with more variables 
lm_test2 = lm(data=data_mod, gcc_90~dayl..s.+tmin..deg.c.)

# predict with remaining years
data_pred = bnh_data %>% 
  dplyr::filter(year>=2014) %>% 
  select(gcc_90, dayl..s., tmin..deg.c., year, doy, date) 

# predict with one var
data_pred$pred1_gcc_90 = predict.lm(lm_test, data_pred)

# predict with 2 vars 
data_pred$pred2_gcc_90 = predict.lm(lm_test2, data_pred)

ggplot(data_pred) + 
  geom_point(aes(x=date, y=pred1_gcc_90, col="1 var")) + 
  geom_point(aes(x=date, y=pred2_gcc_90, col="2 vars"))+ 
  geom_point(aes(x=date, y=gcc_90, col="obs")) 

```

## use models from phenoR package 

Transition dates are 25% of GCC amplitude - included in phenocamr data structure 

Ti = mean daily temps 

Li = daylength 

```{r pars}

# read in parameter ranges 
path <- sprintf("%s/extdata/parameter_ranges.csv",path.package("phenor"))
par_ranges <- read.table(path,
                        header = TRUE,
                        sep = ",")

lin_par = par_ranges %>%
  dplyr::filter(model=="LIN") %>% 
  select(a,b)

# load data from phenor package
bnh = phenocam_DB$bartlettir # in proper format for models 
bnh = phenocam_DB
# plot temps 
tmin = data.frame(bnh$Ti) 
colnames(tmin) <- bnh$year
tmin$doy = bnh$doy
# shifting doy so that 0 is beginning of year jan 1 
tmin$doy = ifelse(tmin$doy<0,
                   tmin$doy+365,
                   tmin$doy)
                  

tmin2 = pivot_longer(tmin, names_to="year", values_to="tmin", cols=-doy)

# plot temperatures 
tmin2 %>% 
ggplot(aes(x=doy, y=tmin, col=year)) + 
  geom_line() 

# make year / spring dates 
spring = data.frame(year=bnh$year, greenup=bnh$transition_dates)

# plot this pattern
ggplot(spring, aes(x=year, y=greenup)) +
  geom_point() +
  geom_smooth(method="lm") 

try_lm = lm(data=spring, greenup~year)

# add dates to plot 
tmin2 %>% 
ggplot(aes(x=doy, y=tmin, col=as.factor(year))) + 
  geom_line() +
  geom_vline(data=spring, aes(xintercept=greenup, col=as.factor(year))) + scale_color_viridis_d()

# plot spring df with tmin 
ann_tmin = tmin2 %>% group_by(year) %>% 
  summarise(tmin_min = min(tmin),
            tmin_max = max(tmin),
            tmin_mn = mean(tmin)) %>% 
  mutate(year=as.double(year)) %>% 
  inner_join(spring, by="year")

ggplot(ann_tmin, aes(x=year, 
                     y=tmin_mn,
                     ymin=tmin_min,
                     ymax=tmin_max)) + geom_pointrange()

ggplot(ann_tmin, aes(y=greenup, 
                     x=tmin_mn)) + 
  geom_point() + geom_smooth(method='lm')

# could also look at spring temps 
# we will define spring as 60-151 (March-Apr-May)
ann_tmin_spr = tmin2 %>% 
  filter(doy>=60 & 
           doy<=151) %>% 
  group_by(year) %>% 
  summarise(tmin_min = min(tmin),
            tmin_max = max(tmin),
            tmin_mn = mean(tmin)) %>% 
  mutate(year=as.double(year)) %>% 
  inner_join(spring, by="year")

# plot with mean min tmps
ggplot(ann_tmin_spr, aes(y=greenup, 
                     x=tmin_mn)) + 
  geom_point() + geom_smooth(method='lm')

# estimate params by working backwards with 
spr_lm = lm(data=ann_tmin_spr, greenup~tmin_mn)
summary(spr_lm)

# data must be formatted for input 
# LIN runs a linear model of spring mean temps 
# doy = a*mean_spring_temp + b 
tmp = LIN(par=c(a=-2.3, b=138.8), data=bnh)

ann_tmin_spr$lm_pred = tmp

ggplot(ann_tmin_spr) +
  geom_point(aes(x=greenup, y=lm_pred)) +
  geom_abline(aes(intercept=0, slope=1)) +
  labs(x="observed", y="predicted")

```
